Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-05-30 document/ms.tex
changing units of plots
incorporating a few more comments from tim

\paragraph{Inverse-detection-efficiency}
It's now interesting to take a brief aside and discuss the connection between
our model and the commonly used inverse-detection-efficiency procedure
(\citealt{howard,dressing,petigura}).
This procedure involves making a weighted histogram of the catalog entries
where the weight for object $\entry_k$ is $1/\completeness(\entry_k)$.
This makes intuitive sense but (to our knowledge) does not have a clear
probabilistic justification or interpretation.

If we model the occurrence rate as a histogram with $J$ fixed bin volumes
$\binarea_j$
\begin{eqnarray}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\ratepar_1 & \entry \in \binarea_1,\\
\ratepar_2 & \entry \in \binarea_2,\\
\cdots \\
\ratepar_J & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
then \eq{poisson-like} becomes
\begin{eqnarray}
\ln p(\{\entry_k\}\,|\,\ratepars) &=&
    \sum_{k=1}^K \sum_{j=1}^J \mathbf{1}[\entry_k \in
        \binarea_j]\,\ln\completeness(\entry_k)\,\ratepar_j
    -\sum_{j=1}^J\ratepar_j\,\int_{\binarea_j} \completeness(\entry)\dd\entry
\end{eqnarray}
where the indicator function $\mathbf{1}[\cdot]$ is one if $\cdot$ is true and
zero otherwise.
If we take the gradient of this function with respect to \ratepars\ and set it
equal to zero, we find the maximum likelihood result
\begin{eqnarray}
\ratepar_j &=& \frac{N_j}{\int_{\binarea_j} \completeness(\entry)\dd\entry}
\end{eqnarray}
where $N_j$ is the number of objects that fall within the bin $j$.
Since it is the maximum likelihood solution, this result is the minimum
variance estimator for the occurrence rate in the limit of negligible catalog
uncertainties.
This is \emph{almost} the same as the inverse-detection-efficiency result and
in various limiting cases, they would provide the same result.
\dfmtodo{Finish explaining why this result is better under the specified
assumptions.}

\paragraph{occurrence rate parameterization \& priors}
For the remainder of this \paper, we also model the occurrence rate as a
two-dimensional histogram with fixed logarithmic bins in period and radius.
When we include observational uncertainties---using \eq{money}---the maximum
likelihood result is no longer analytic.
Therefore, if we want to compute the ``best-fit'' occurrence rate, we can use a
standard non-linear optimization algorithm.
In the regions of parameter space that we tend to care about, the completeness
is low and there are only a few observations with large uncertainties.
In this case, we're especially interested in probabilistic constraints on the
occurrence rate; not just the best-fit model.
To do this, we apply a prior\footnote{This is a slight misnomer because
we also we also marginalize over the parameters of this model.} $p(\ratepars)$
on the occurrence rate parameters and generate samples from the posterior
probability
\begin{eqnarray}\eqlabel{posterior}
p(\ratepars\,|\,\{\data_k\}) &\propto&
    p(\ratepars)\,p(\{\data_k\}\,|\,\ratepars)
\end{eqnarray}
using Markov chain Monte Carlo (MCMC).

@@

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/results/period.pdf}
\end{center}
\caption{%
The period distribution.
\figlabel{period}}
\end{figure}

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/results/radius.pdf}
\end{center}
\caption{%
The radius distribution.
\figlabel{radius}}
\end{figure}

>>>>

\paragraph{Inverse-detection-efficiency}
It's now interesting to take a brief aside and discuss the connection between
our model and the commonly used inverse-detection-efficiency procedure
(\citealt{howard,dressing,petigura}).
This procedure involves making a weighted histogram of the catalog entries
where the weight for object $\entry_k$ is $1/\completeness(\entry_k)$.
This makes intuitive sense but the standard formulation does not have a clear
probabilistic justification or interpretation.
\hoggtodo{Find the papers that Tremaine suggested that we look at re:vmax and
discuss the motivation in those cases.}

If we model the occurrence rate as a histogram with $J$ fixed bin volumes
$\binarea_j$
\begin{eqnarray}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\ratepar_1 & \entry \in \binarea_1,\\
\ratepar_2 & \entry \in \binarea_2,\\
\cdots \\
\ratepar_J & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
then \eq{poisson-like} becomes
\begin{eqnarray}
\ln p(\{\entry_k\}\,|\,\ratepars) &=&
    \sum_{k=1}^K \sum_{j=1}^J \mathbf{1}[\entry_k \in
        \binarea_j]\,\ln\completeness(\entry_k)\,\ratepar_j
    -\sum_{j=1}^J\ratepar_j\,\int_{\binarea_j} \completeness(\entry)\dd\entry
\end{eqnarray}
where the indicator function $\mathbf{1}[\cdot]$ is one if $\cdot$ is true and
zero otherwise.
If we take the gradient of this function with respect to \ratepars\ and set it
equal to zero, we find the maximum likelihood result
\begin{eqnarray}\eqlabel{ml-ide}
{\ratepar_j}^* &=&
\frac{N_j}{\int_{\binarea_j} \completeness(\entry)\dd\entry}
\end{eqnarray}
where $N_j$ is the number of objects that fall within the bin $j$.
We estimate the uncertainty $\delta\ratepar_j$ on this value by examining the
curvature of the log-likelihood function near the maximum and find
\begin{eqnarray}
\frac{\delta\ratepar_j}{{\ratepar_j}^*} &=& \frac{1}{\sqrt{N_j}} \quad.
\end{eqnarray}

Since \eq{ml-ide} gives the maximum likelihood solution, this result is the
minimum-variance estimator for the occurrence rate when the completeness
function $\completeness(\entry)$ is known and the catalog uncertainties are
negligible.
This is \emph{almost} the same as the inverse-detection-efficiency procedure
and if the completeness function is constant in each bin \binarea, both
methods give the same result.
If the completeness function is not piecewise-constant on the same grid as the
occurrence rate inference, the inverse-detection-efficiency procedure will
give a higher variance estimate of the occurrence rate.

\paragraph{Occurrence rate parameterization \& priors}
For the remainder of this \paper, we also model the occurrence rate as a
two-dimensional histogram with fixed logarithmic bins in period and radius.
When we include observational uncertainties---using \eq{money}---the maximum
likelihood result is no longer analytic.
Therefore, if we want to compute the ``best-fit'' occurrence rate, we can use a
standard non-linear optimization algorithm.
In the regions of parameter space that we tend to care about, the completeness
is low and there are only a few observations with large uncertainties.
In this case, we're especially interested in probabilistic constraints on the
occurrence rate; not just the best-fit model.
To do this, we apply a prior\footnote{This is a slight misnomer because
we also we also marginalize over the parameters of this model.} $p(\ratepars)$
on the occurrence rate parameters and generate samples from the posterior
probability
\begin{eqnarray}\eqlabel{posterior}
p(\ratepars\,|\,\{\data_k\}) &\propto&
    p(\ratepars)\,p(\{\data_k\}\,|\,\ratepars)
\end{eqnarray}
using Markov chain Monte Carlo (MCMC).

@@

\begin{figure}[p]
\begin{center}
\includegraphics{figures/results/period.pdf}
\end{center}
\caption{%
The period distribution.
\figlabel{period}}
\end{figure}

\begin{figure}[p]
\begin{center}
\includegraphics{figures/results/radius.pdf}
\end{center}
\caption{%
The radius distribution.
\figlabel{radius}}
\end{figure}

\begin{figure}[p]
\begin{center}
\includegraphics{figures/results/linear-radius.pdf}
\end{center}
\caption{%
The radius distribution.
\figlabel{linear-radius}}
\end{figure}
